Appendix B: The Forecast Trap in a Sequential Decision-Making Problem

### 2022-04-12 manuscript/appendix-A.Rmd
:pen:
here we go

This appendix illustrates how the forecast trap can emerge in choosing between two alternative three-species models which may be used to manage a partially observed five-species system. 
The underlying dynamics for the simulated "true" system are based on the five-dimensional model of a herring fishery presented in @Brias2021.
Following @Brias2021, we assume three species of herring, $X_1$, $X_2$, $X_3$ are preyed upon by

\begin{align}
x_{1,t+1} &= s_{1,t} \exp{(1.0213 -        s_{1,t} - 0.0861 s_{2,t} - 0.3141 s_{3,t} - 0.7252 s_{4,t} -0.2445 x_{5,t} +\espilon_{1,t} )}
x_{2,t+1} &= s_{2,t} \exp{(1.0289 - 0.4765 s_{1,t} -        s_{2,t} - 0.1370 s_{3,t} - 0.9811 s_{4,t} -0.0915 x_{5,t} +\espilon_{2,t} )}
x_{3,t+1} &= s_{3,t} \exp{(1.0207 - 0.3193 s_{1,t} - 0.3461 s_{2,t} -        s_{3,t} - 0.6367 s_{4,t} -0.8716 x_{5,t} +\espilon_{3,t} )}
x_{4,t+1} &= s_{4,t} \exp{(         0.7252 s_{1,t} + 0.9811 s_{2,t} + 0.6367 s_{3,t} -        s_{4,t}                 +\espilon_{4,t})}
x_{5,t+1} &= x_{5,t} \exp{(         0.2445 s_{1,t} + 0.0915 s {2,t} + 0.8716 s {3,t} -                        x_{5,t} +\espilon_{5,t} )}
\end{align}

@@

\begin{equation}
s_{i,t} = x_{i_t} \left(1 - u_{i,t}\right)
\end{equation}

\begin{equation}
\epsilon_{i,t} \stackrel{\texrm{iid}}{\tilde} \mathcal{N(0, \sigma_i)}
\end{equation}

@@

Model A is a possible 3-species approximation that produces more accurate forecasts but leads to substantially worse management outcomes
This model treats the three species of herring collectively as $x_1$. 
Parameters are estimated consistent with limited historical data of the observed variables, bass biomass $x_4$ and herring biomass $x_5$.

\begin{equation}
x_{1,t+1} &= s_{1,t} \exp(3 - 8 s_{1,t} - 0.9 s_4 - 0.9 * x_{5,t} + \espilon_{1,t} )
x_{4,t+1} &= s_{4,t} \exp(0.1 s_{1,t} - s_{4,t} + \espilon_{4,t}  )
x_{5,t+1} &= x_{5,t} \exp(0.3 s_{1,t} - x_5  + \espilon_{5,t}  )
\end{equation}

@@

\begin{equation}
x_{1,t+1} &= s_{1,t} \exp(1.60 - 2.00 s_{1,t}  + \espilon_{1,t} )
x_{4,t+1} &= s_{4,t} \exp(1.45 - 2.75 s_{1,t}  + \espilon_{4,t}  )
x_{5,t+1} &= 0.25 x_{5,t} 
\end{equation}

@@

knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)

@@

```{r}
model_B <- function(u_prey = 0,
                      u_predator = 0,
                      A1 = 2,
                      R1 = 1.6,
                      A2 = 2.75,
                      R2 = 1.45,
                      Tmax = 50,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      sigma = 0.01,
                      reps=40) {
    df <- tibble()
    for(r in 1:reps){
    omega1 <- rnorm(Tmax, 0, sigma)
    omega2 <- rnorm(Tmax, 0, sigma)
    omega3 <- rnorm(Tmax, 0, sigma)

@@

    for (t in 1:(Tmax-1)) {
        s1[t] <- x1[t] * (1 - 10*u_prey)
        s4[t] <- x4[t] * (1 - u_predator)

@@

```{r}
u_A <- function(x) utility(x, f = model_A)
A <- optim(c(.2,0.01), u_A)
model_A_policy <- clip(A$par, 0, 1)
```

```{r}
u_B <- function(x) utility(x, f = model_B)
B <- optim(c(0.5,0.1), u_B)
model_B_policy <- clip(B$par, 0, 1)
```

@@

```{r}
mean_utility_A / mean_utility_B
```

>>>>

This appendix illustrates how the forecast trap can emerge in choosing between two alternative three-species models which may be used to manage a partially observed five-species system. 
The underlying dynamics for the simulated "true" system are based on the five-dimensional model of a herring fishery presented in @Brias2021.
I have used the model parameters from @Brias2021, which were selected to ensure co-existence of the five species as outlined in their paper.
Following @Brias2021, we assume three species of herring, $X_1$, $X_2$, $X_3$ are preyed upon by

\begin{align}
x_{1,t+1} &= s_{1,t} \exp{(1.0213 -        s_{1,t} - 0.0861 s_{2,t} - 0.3141 s_{3,t} - 0.7252 s_{4,t} -0.2445 x_{5,t} +\epsilon_{1,t} )} \\
x_{2,t+1} &= s_{2,t} \exp{(1.0289 - 0.4765 s_{1,t} -        s_{2,t} - 0.1370 s_{3,t} - 0.9811 s_{4,t} -0.0915 x_{5,t} +\epsilon_{2,t} )} \\
x_{3,t+1} &= s_{3,t} \exp{(1.0207 - 0.3193 s_{1,t} - 0.3461 s_{2,t} -        s_{3,t} - 0.6367 s_{4,t} -0.8716 x_{5,t} +\epsilon_{3,t} )} \\
x_{4,t+1} &= s_{4,t} \exp{(         0.7252 s_{1,t} + 0.9811 s_{2,t} + 0.6367 s_{3,t} -        s_{4,t}                 +\epsilon_{4,t})} \\
x_{5,t+1} &= x_{5,t} \exp{(         0.2445 s_{1,t} + 0.0915 s {2,t} + 0.8716 s {3,t} -                        x_{5,t} +\epsilon_{5,t} )}
\end{align}

@@

$$
s_{i,t} = x_{i_t} \left( 1 - u_{i,t} \right)
$$

$$
\epsilon_{i,t} \stackrel{iid}{\sim} \mathcal{N}(0, \sigma_i)
$$

@@

Model A is a possible 3-species approximation that produces more accurate forecasts but leads to substantially worse management outcomes
This model treats the three species of herring collectively as $x_1$. 
As with the original @Brias2021 model, parameters are not based directly on historical fisheries data, but have been selected to be consistent with simulations of the observed variables from the "true" model of @Brias2021, bass biomass $x_4$ and herring biomass $x_5$.

\begin{align}
x_{1,t+1} &= s_{1,t} \exp(3 - 8 s_{1,t} - 0.9 s_4 - 0.9 * x_{5,t} + \epsilon_{1,t} ) \\
x_{4,t+1} &= s_{4,t} \exp(0.1 s_{1,t} - s_{4,t} + \epsilon_{4,t}  ) \\
x_{5,t+1} &= x_{5,t} \exp(0.3 s_{1,t} - x_5  + \epsilon_{5,t}  )
\end{align}

@@

\begin{align}
x_{1,t+1} &= s_{1,t} \exp(0.1 - 2.00 s_{1,t}  + \epsilon_{1,t} ) \\
x_{4,t+1} &= s_{4,t} \exp(1.45 - 2.75 s_{1,t}  + \epsilon_{4,t}  ) \\
x_{5,t+1} &= 0.25 x_{5,t} 
\end{align}

@@

knitr::opts_chunk$set(cache=TRUE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)

set.seed(123)
```

@@

```{r}
model_B <- function(u_prey = 0,
                      u_predator = 0,
                      A1 = 2,
                      R1 = 0.1,
                      A2 = 2.75,
                      R2 = 1.45,
                      Tmax = 50,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      sigma = 0.01,
                      reps=40) {
    df <- tibble()
    for(r in 1:reps){
    omega1 <- rnorm(Tmax, 0, sigma)
    omega2 <- rnorm(Tmax, 0, sigma)
    omega3 <- rnorm(Tmax, 0, sigma)

@@

    for (t in 1:(Tmax-1)) {
        s1[t] <- x1[t] * (1 - u_prey)
        s4[t] <- x4[t] * (1 - u_predator)

@@

```{r}
u_A <- function(x) utility(x, f = model_A)
A <- optim(c(.8,0.02), u_A)
model_A_policy <- clip(A$par, 0, 1)
```

```{r}
u_B <- function(x) utility(x, f = model_B)
B <- optim(c(0.01,0.1), u_B)
model_B_policy <- clip(B$par, 0, 1)
```

@@

```{r}
ratio <- mean_utility_A / mean_utility_B
```

The ratio of mean utility under model A to mean utility under the nearly-optimal model B is `r round(ratio *100)`%.

# References

### 2022-04-12 manuscript/appendix-B.Rmd
here we go

Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_t+1$, and we know the current stock, $X_t$, then we can sustainably harvest $X_{t+1} - X_t$ without decreasing the biomass over the long term.
Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of overfitting by comparing model predictions to later observations that were not used to estimate the model [@Gneiting2014].

@@

We use value iteration to solve the stochastic dynamic program [@Marescot2013; @MDPtoolbox] for each model.
This determines the optimal harvest policy for each possible state, given each model.
Because this step is the most computationally intensive routine, we cache the results using memosization conditioned on the transition matrices [@memoise].
Running this code with alternate transition matrices automatically invalidates that cache, reducing the risk of loading spurious results.

@@

am_npv <- sum(am$value * discount ^ am$time)
am_economics_percent <- round(am_npv / npv[[1,"npv"]] * 100)

mean_state <- sims %>% group_by(model) %>% summarize(state = mean(state))
am_ecology_percent <- round(mean(states[am$state])/mean_state[[1, "state"]]*100)
```

>>>>

Like many decision problems, the task of setting a sustainable harvest quota appears to hinge on having an accurate forecast: if we can predict to what size the fish stock will increase next year, $X_{t+1}$, and we know the current stock, $X_t$, then we can sustainably harvest $X_{t+1} - X_t$ without decreasing the biomass over the long term.
Selecting a model based on forecast skill is also justifiable on theoretical grounds, since it reduces the risk of overfitting by comparing model predictions to later observations that were not used to estimate the model [@Gneiting2014].

@@

We use value iteration to solve the stochastic dynamic program [@Marescot2013; @MDPtoolbox] for each model.
This determines the optimal harvest policy for each possible state, given each model.
Because this step is the most computationally intensive routine, we cache the results using memoization conditioned on the transition matrices [@memoise].
Running this code with alternate transition matrices automatically invalidates that cache, reducing the risk of loading spurious results.

@@

am_npv <- am %>% group_by(method) %>% summarize(npv = sum(quota * discount ^ time))
learning <- am_npv %>% filter(method=="learning") %>% pull(npv) 
am_economics_percent <- round(learning / npv[[1,"npv"]] * 100)

mean_state <- sims %>%
  group_by(model) %>% 
  summarize(state = mean(state)) %>% 
  filter(model == 1) %>%
  pull(state)
learn_state <- am %>%
  group_by(method) %>% 
  summarize(state = mean(stock))  %>%
  filter(method == "learning") %>%
  pull(state)

am_ecology_percent <- round(learn_state / mean_state *100)
```
